Global code motion of congruent instructions

### 2017-04-17 paper.tex
change algos

This paper describes code-motion of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting can reduce the critical path length of execution in
out-of-order machines. As more instructions are available at the hoisting point,
the hardware has more independent instructions to reorder. The following example
illustrates how hoisting can improve performance by exposing more ILP.

@@

In fact there are several advantages of code-motion of this kind:
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
however, a general implementation of code-motion of congruent (non redundant)
expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size and improve performance in some
  cases.
\end{itemize}

@@

Barany \cite{barany2013} presented a global scheduler with integer linear
programming (ILP) formulation with a goal to minimize register pressure. The
results they got were not very promising. It may be because they only used the
scheduler for smaller functions ($<1000$ instructions); also, they compiled the
benchmarks for ARM-Cortex which is more resilient to register pressure because
it has more registers compared to X86, for example.

@@

With the algorithm described in the next section, we got a
reduction in register spills, a reduction in code-size and some performance
improvements with very low compilation time overhead, on \SPEC{} benchmarks.

@@

The code-motion pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for code-motion,
\item compute a point in the program where it is both legal and profitable to
  move the code,
\item move the code to hoist point or sink point, and
\item update data structures to continue iterative code motion.
\end{itemize}

@@

We compute merge sets of the control flow graph which remains the same through
the code-motion transformation. For code-motion we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of isLiveOutUsingMergeSet is
presented here:

@@

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation. If a variable is not live-out of a basic block, it
still may be used later in the basic block: to establish whether a use of a
variable is a kill we iterate on all the subsequent instructions in a basic
block checking for uses.

@@

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
B1 and those in B5, B6 to be hoisted at B2. Even if (B3, B4) could not be
hoisted, for some reason, to B1, (B5, B6) can still be hoisted -- if legal and
profitable -- to B2.

@@

Code-motion basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining.) The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

@@

% Explain the overall algorithm, explain how anticipability is computed on demand
\begin{verbatim}
Analyses available: Dominator Tree, DFS Numbering, Memory SSA,

@@

If any of candidates were hoisted then repeat hoisting
Once no more candidates are hoistable proceed to sinking
\end{verbatim}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
Benchmark       &\% change in performance \% change in codesize & \\\hline
400.perlbench	& 1.00  & 1.00 \\\hline
401.bzip2	& 1.00  & 1.00 \\\hline
403.gcc	        & 1.00  & 1.01 \\\hline
429.mcf	        & 1.03  & 1.00 \\\hline
433.milc	& 1.00  & 1.00 \\\hline
444.namd	& 1.00  & 1.01 \\\hline
445.gobmk	& 0.99  & 1.01 \\\hline
447.dealII	& 0.95  & 1.01 \\\hline
450.soplex	& 0.99  & 0.99 \\\hline
453.povray	& 0.97  & 1.00 \\\hline
456.hmmer	& 0.99  & 1.00 \\\hline
458.sjeng	& 1.02  & 1.00 \\\hline
462.libquantum	& 1.00  & 1.00 \\\hline
464.h264ref	& 1.00  & 1.00 \\\hline
470.lbm	        & 1.00  & 1.00 \\\hline
471.omnetpp	& 1.00  & 0.99 \\\hline
473.astar	& 1.00  & 1.02 \\\hline
482.sphinx3	& 0.99  & 1.00 \\\hline
483.xalancbmk	& 1.01  & 0.99 \\\hline
Geomean         & 0.997 & 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:code-size}
\end{table}

>>>>

This paper describes code-motion of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting, for example, can reduce the critical path length of
execution in out-of-order machines. As more instructions are available at the
hoisting point, the hardware has more independent instructions to reorder. The
following example illustrates how hoisting can improve performance by exposing
more ILP.

@@

In fact, there are several advantages of GCM.
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP), which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored code-motion of congruent instructions in as much detail as we
have done. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk, however, a general implementation of code-motion of
congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
trunk. The main contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

Barany \cite{barany2013} presented a global scheduler with ILP formulation with
a goal to minimize register pressure. The results they got were not very
promising. It may be because they only used the scheduler for smaller functions
($<1000$ instructions); also, they compiled the benchmarks for ARM-Cortex which
is more resilient to register pressure because it has more registers compared to
X86, for example.

@@

With the algorithm described in the next section, we got a reduction in register
spills, improved inliner heuristics, improved compile time and some performance
improvements on \SPEC{} benchmarks.

@@

The GCM pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for code-motion,
\item compute a point in the program where it is both legal and profitable to
  move the code,
\item move the code to hoist point or sink point, and
\item update data structures to continue iterative code motion.
\end{itemize}

@@

We compute merge sets of the control flow graph which remains the same through
the code-motion transformation. For code-motion we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation. If a variable is not live-out of a basic block, it
still may be used later in the basic block: to establish whether a use of a
variable is a kill we iterate on all the subsequent instructions in a basic
block checking for uses, the \emph{isKill} function uses
\emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
instruction.

@@

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted -- if legal and profitable -- to B2.

@@

Code-motion basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

@@

% Explain the overall algorithm, explain how anticipability is computed on demand
\begin{verbatim}
int doCodeHoisting()
  Analyses available: Dominator Tree, DFS Numbering, Memory SSA,

@@

  return number_of_hoisted_candidates

If any of candidates were hoisted then repeat hoisting
\end{verbatim}

Once no more candidates are hoistable proceed to sinking

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
Benchmark       &\% change in performance & \% change in codesize \\\hline
400.perlbench	& 1.00  & 1.00 \\\hline
401.bzip2	& 1.00  & 1.00 \\\hline
403.gcc	        & 1.00  & 1.01 \\\hline
429.mcf	        & 1.03  & 1.00 \\\hline
433.milc	& 1.00  & 1.00 \\\hline
444.namd	& 1.00  & 1.01 \\\hline
445.gobmk	& 0.99  & 1.01 \\\hline
447.dealII	& 0.95  & 1.01 \\\hline
450.soplex	& 0.99  & 0.99 \\\hline
453.povray	& 0.97  & 1.00 \\\hline
456.hmmer	& 0.99  & 1.00 \\\hline
458.sjeng	& 1.02  & 1.00 \\\hline
462.libquantum	& 1.00  & 1.00 \\\hline
464.h264ref	& 1.00  & 1.00 \\\hline
470.lbm	        & 1.00  & 1.00 \\\hline
471.omnetpp	& 1.00  & 0.99 \\\hline
473.astar	& 1.00  & 1.02 \\\hline
482.sphinx3	& 0.99  & 1.00 \\\hline
483.xalancbmk	& 1.01  & 0.99 \\\hline
Geomean         & 0.997 & 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:code-size}
\end{table}
